Python as a machine learning language: experiences from the scikit-learn project

### 2013-05-01 paper.tex
name/affilition for current authors (per git shortlog :)
@mblondel, @jakevdp, please check and correct.

I've sorted author names alphabetically; let's argue about the order
later.
some problems with SciPy and Cython
Hope I hedged this carefully enough.

@@

>>>>

% Add your name and affiliation here.
% Don't worry too much about the formatting,
% as we'll have to redo that for the LNCS format.
\author{Mathieu Blondel \\
        Graduate School of System Informatics \\
        Kobe University \\
        \small{\texttt{mblondel@ai.cs.kobe-u.ac.jp}}
      \and
        Lars Buitinck \\
        Informatics Institute \\
        University of Amsterdam \\
        \small{\texttt{L.J.Buitinck@uva.nl}}
      \and
        Jake Vanderplas \\
        Astronomy Department \\
        University of Washington \\
        \small{\texttt{vanderplas@astro.washington.edu}}
}

@@

\section{Weaknesses of the NumPy/SciPy environment}

% This needs an introductory paragraph, but the best I could come up with was:
As can be expected, the NumPy/SciPy environment is not completely flawless.

% The following hedge needs to be in here; we don't want to diss our peers
It should be noted that this section is meant
as a summary of challenges faced by scikit-learn development
and an attempt at constructive criticism.
There is considerable overlap between the communities
that develop or contribute to NumPy, SciPy, scikit-learn, and Cython,
so the issues raised here may eventually be resolved
by the scikit-learn developers themselves.

\subsection{Sparse matrix support}

SciPy's handling of sparse matrices is one source of confusion
for scikit-learn developers and users alike.
The \texttt{scipy.sparse} package offers quite full-featured support
for this kind of data structure,
which is crucial to achieving performance and scalability
in the face of high-dimensional inputs
(including, but not limited to, NLP applications).
However, the interface exposed by its classes
deviates from the NumPy array interface, sometimes subtly so,
sometimes entirely.  % XXX is this grammatically correct?
For example, while NumPy arrays have a variable number of dimensions,
so that vectors can be considered 1-d arrays
(without a distinction between row and column vectors),
matrices are 2-d, etc., \texttt{scipy.sparse} matrices are always 2-dimensional,
meaning that special care has to be taken when handling sparse vectors.

The development team has built up sufficient familiarity
with the sparse/array distinction to factor out common patterns
for supporting both sparse and dense input data structures in many estimators.
In other cases, though, sparse matrix support has to implemented
on a per-estimator basis.
This situation is aggravated by the lack of a C API for sparse matrices,
so that Cython modules have to implement their own version
of operations like sparse dot products
that ideally would have been handled by a library.

However, users of scikit-learn cannot always be expected
to have similar familiarity, and in some cases
(e.g., when users want to leverage feature extraction code
outside the context of a \texttt{Pipeline})
they will have to face the \texttt{scipy.sparse} interface.
This is, of course, as much an artifact of the design choice
to not implement separate classes and interfaces for samples/instances
as it a consequence of SciPy's implementation of sparse matrices.

\subsection{Python/Cython mismatch}

The Python/Cython divide is another source of potential difficulties
both for users and developers/contributors.
While Cython makes writing performance-critical code possible
in a language that is very much like Python,
doing so properly still requires some knowledge of C.
One source of trouble has been the various fixed-size integer types
that are used by the Python and NumPy C runtimes,
whose sizes are practically all platform-dependent.
Choosing the appropriate integer type from among
\texttt{int}, \texttt{size\_t}, \texttt{Py\_ssize\_t} and \texttt{npy\_intp}
can require knowledge of the Python extension API,
the NumPy C API and the C standard.
Neither Cython nor C compilers provide much in the way of static checking
for the right type,
while picking the wrong type may cause segmentation faults
when users try to use arrays that are too large for the type used
and using a dynamic type often leads to unacceptable slowdowns.

Another problem concerns the mismatch between on the one hand Python functions,
which are first-class runtime objects endowed with methods and metadata
that identify them as ``callable'',
that dynamically check their number and types of arguments,
and that can return any Python type of object,
and compiled C or Cython functions on the other hand,
which exhibit static typing.
Where it may sometimes be convenient for estimators
to take functions as arguments
(e.g., in the case of kernel machine learning,
where one might like to use a custom kernel),
but such usage is not at present possible and may never be.
Passing C or Cython-compiled functions would require an API for the kernel method
to be implemented at the C or Cython level,
only to be used from compiled code.
Passing arbitrary Python functions would be possible,
but performance can be expected to degrade significantly
because of the runtime checks needed for each invocation of the kernel function.
(The current solution in scikit-learn is to let the user pass not a kernel function,
but a precomputed Gram matrix for an entire set of objects.
The linear, polynomial and RBF kernels are treated as special cases in
the \textsf{LIBSVM}-based SVM learner, where they are implemented in C{}\verb!++!.)
